### 2020-07-01 manuscript/manuscript.rmd
update supplementary orders in the manuscript and add codes to generate them in kernel

All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the Ctenophora-sister hypothesis (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 3), the condiitions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

@@

A primary difference between site-heterogeneous models with C (like C60) and CAT are the number of substitutional categories that sites are assigned to. CAT models employ a Dirichlet process prior to infer number of substitutional categories across a phylogenetic data matrix, so the number of categories is variable. Models with C60 are set to 60 categories. Differences in analysis results could therefore be due to differences in the number of categories. We therefore first assessed the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 5). (XX add mean number of ) The number of categories is considerable larger than 60 categories used in C60 models, with a mean of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This is a very large number of additional parameters that must be estimated from the data. When ModelFinder is used to select the number of categories in IG-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria. The fact that CAT is selecting ten fold more than 60 is consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 7). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of categories. When we sweep the number of categories, we can observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4), all within PhyloBayes. For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

@@

ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")), nrow = 3,
          p3, p4, labels = c("C", "D"))
```
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values are from the sensitive analyses are shown in Supplementary Table 7.

@@

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 4). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

@@

Supplementary_Table 4. Distribution of phylogenetic signal of differnet models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).

Supplementary_Table 5. Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).

Supplementary_Table 6. Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).

Supplementary_Table 7. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices ((Table is converted from analyses_sensitive in Rdata)).

>>>>

All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the Ctenophora-sister hypothesis (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

@@

A primary difference between site-heterogeneous models with C (like C60) and CAT are the number of substitutional categories that sites are assigned to. CAT models employ a Dirichlet process prior to infer number of substitutional categories across a phylogenetic data matrix, so the number of categories is variable. Models with C60 are set to 60 categories. Differences in analysis results could therefore be due to differences in the number of categories. We therefore first assessed the number of categories inferred from CAT model for each PhyloBayes analysis (Supplementary Table 4). (XX add mean number of ) The number of categories is considerable larger than 60 categories used in C60 models, with a mean of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This is a very large number of additional parameters that must be estimated from the data. When ModelFinder is used to select the number of categories in IG-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria (Supplementary Table 5). The fact that CAT is selecting ten fold more than 60 is consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of categories. When we sweep the number of categories, we can observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4), all within PhyloBayes. For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

@@

ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")), nrow = 3,
          p3, p4, labels = c("C", "D"))
```
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values are from the sensitive analyses are shown in Supplementary Table 6.

@@

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 7). These results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

@@

Supplementary_Table 4. Summary statistics of CAT substitutional categories inferred from different matrices (Table is manualy curated from Tracer result for each PhyloBayes analysis).

Supplementary_Table 5. Summary of amino acid frequencies of 60 categories inferred by C60 model using IQ-TREE (Table is manualy curated from IQtree log file for each analysis).

Supplementary_Table 6. Summary of sensitive analyses with different number of CAT substitutional categories in representative matrices ((Table is converted from analyses_sensitive in Rdata)).

Supplementary_Table 7. Distribution of phylogenetic signal of different models and outgroup sampling for two alternative hypotheses on the animal-root position in Philippe2009, Ryan2013 and Whelan2017_full matrices (Table is converted from au_tests in Rdata).
